Explicit B-spline regularization in diffeomorphic image registration

### 2013-10-28 paper.tex
WIP: minor ba edits

\begin{document}
\onecolumn
\firstpage{1}

\title[DMFFD diffeomorphic image registration]{Explicit B-spline regularization in diffeomorphic image registration}
\author[\firstAuthorLast ]{\Authors}
\address{}
\correspondance{}
\editor{}
\topic{}

@@

\begin{abstract}
Important methodological developments in the evolution of image registration algorithms include those
in which the correspondence relationship is characterized by diffeomorphisms.
The  popularity of these approaches is due largely to their
topological properties and success in providing biologically plausible
solutions to deformation and morphological
estimation problems. Popular variants of diffeomorphic algorithms include those characterized by
time-varying and constant velocity fields, and symmetrical considerations.
Prior information in the form of regularization is used to enforce transform plausibility taking the form of physics-based constraints or through some approximation thereof, e.g. Gaussian smoothing of the vector fields (a la Thirion's Demons \citep{thirion1998}).  In the context of the original Demons' framework, the {\it directly manipulated free-form deformation} approach \citep{tustison2009}, can be viewed as a smoothing alternative in which explicit regularization is achieved through fast B-spline approximation.
This characterization can be used to provide B-spline ``flavored'' diffeomorphic image registration solutions with several advantages.  Implementation is open source and available through the Insight Toolkit and our Advanced Normalization Tools (ANTs) repository.  A thorough comparative evaluation with the well-known SyN algorithm \citep{avants2008}, implemented within the same framework, and its B-spline analog is performed using open labeled brain data and open source evaluation tools.  
\tiny
\keyFont{ \section{Keywords:}
Advanced Normalization Tools, diffeomorphisms, directly manipulated free-form deformation, Insight Toolkit, spatial normalization }
\end{abstract}

@@

Establishment of anatomical and functional correspondence
is a crucial step towards gaining insight into certain biological
sciences.  Neuroscience research efforts, such as characterizing
brain morphology, require accurate and robust methods for
producing such mappings.  The
extensive literature detailing methodology is evidence of the rich history of
algorithmic development which continues contemporaneously.
We highlight several key historical contributions which are particularly relevant to the work 
subsequently presented.

Free-form deformation (FFD) image registration, characterized by  regularization based on the B-spline basis functions, has several
advantages including algorithmic simplicity,
good performance, desirable properties, and several available
implementations.  Current research was
preceded by related work for
geometric modeling \citep{sederberg1986} and originated with such important
contributions as \cite{szeliski1997,thevenaz1998,rueckert1999}.
Continued development within this early spline-based paradigm produced additional innovations such as integrated similarity metrics \citep[e.g.][]{mattes2003}, additional transformation constraints \citep[e.g.][]{rohlfing2003}, and notable open source implementations \citep[e.g.][]{ibanez2005,klein2010,modat2010,shackleford2010}.

@@

Continuing from the work of \cite{christensen1996} and subsequent exploration into the mathematical formalisms of diffeomorphisms \citep[e.g.][]{dupuis1998},
the well-known Large Deformation Diffeomorphic Metric Mapping (LDDMM) algorithm
was proposed in \cite{beg2005}.  In contrast to the mapping produced
by \cite{christensen1996}, LDDMM yields the geodesic solution in the space of diffeomorphisms between two images. Since its introduction, LDDMM has
inspired much innovation in the image registration community.  Applying
the log-Euclidean framework of \cite{arsigny2006}, DARTEL (Diffeomorphic Anatomic Registration using Exponential Lie algebra) uses a constant velocity field parameterization to provide a fast, diffeomorphic alternative \citep{ashburner2007}. Additionally, symmetrical considerations in the velocity field parameterization are discussed in \cite{avants2008} in the context of a cross correlation similarity metric.
By explicitly symmetrizing the LDDMM formulation, this Symmetric Normalization (SyN) minimizes the bias of the resulting transformation when selecting the ``fixed'' and ``moving'' images.  A greedy version of this algorithm has proven successful in neuroimaging \citep{klein2009} and pulmonary \citep{murphy2011} applications.

@@

Variants of three popular diffeomorphic algorithms and their DMFFD analogs
(LDDMM,
DARTEL,
 and SyN)
were implemented by the authors
as part of the recent refactoring of the open source
Insight Toolkit (ITK) although related work had been
previously implemented within the popular Advanced
Normalization Tools (ANTs).%
\footnote{
http://stnava.github.io/ANTs
}
Given the popularity and excellent performance of the greedy variant of SyN algorithm, our evaluation focus in this work is
its B-spline analog which we denote as ``B-spline SyN'' or ``DMFFD SyN.''
Evaluation and comparisons of the respective algorithmic instantiations
are performed using the \verb#antsRegistration# program found
in the ANTs repository (also originally developed by the authors).  This permits a direct algorithmic comparison
as potential sources for implementation bias have been removed \citep{tustison2013}.  Additionally, in the spirit of open science,
all text, figures, and
scripts to reproduce the results contained in this work are publicly available online.%
\footnote{
https://github.com/ntustison/BSplineMorphisms
}

@@

However,
as pointed out in \cite{avants2008}, implementations of this standard
LDDMM formulation are negatively affected by the lack of optimization
symmetry where arbitrary assignment of fixed and moving images could
lead to different solutions despite the fact that the theoretical
geodesic solution describes the same parameterized path forwards and backwards.
This observation led to the symmetric
formulation of Eq. (\ref{eq:lddmm}) found in \cite{avants2008}:
\begin{align}
  \inf_{\phi_1} \inf_{\phi_2} \left(
                     \int_0^{0.5} \left( \|v_1(t)\|_L^2 + \|v_2(t)\|_L^2 \right) dt +
                     \int_{\Omega} | I \circ \phi_1^{-1}(\mathbf{x},0.5)
                           - J \circ \phi_2^{-1}(\mathbf{x},0.5) |^2 d\Omega
              \right)
\end{align}
where
\begin{align}
  \frac{d \phi_i(\mathbf{x},t)}{dt} = v_i( \phi_i(\mathbf{x},t), t ),\,\, \phi_i(\mathbf{x},0) = \mathbf{Id}, \,\, i \in \{1,2\}.
\end{align}
With extension to arbitrary similarity metric choice, the second
term is replaced with
\begin{align}
\int_{\Omega} \Pi_{\sim}
                          \left( I \circ \phi_1^{-1}(\mathbf{x},0.5),
                           J \circ \phi_2^{-1}(\mathbf{x},0.5) \right) d\Omega
\end{align}
with a popular choice for $\Pi_{\sim}$ being a local neighborhood cross
correlation \citep{avants2008,avants2011}.
Note that
$t$ is parameterized in opposite directions between $\phi_1$ and $\phi_2$.
A diagrammatic
illustration of the explicit symmetry associated with SyN is shown in
Figure \ref{fig:syn}.

\begin{figure}[htb]
  \centering
  \includegraphics[width=0.95\textwidth]{SyN.jpg}
  \caption{Illustration of the greedy SyN formulation.  Given images
  $I_A$ and $I_B$, the symmetric set-up requires finding the two
  transform pairs $\left(\phi_1,\phi_1^{-1}\right)$
  $\left(\phi_2,\phi_2^{-1}\right)$ which map to/from
  the respective images to the midway point. During
  optimization, the update field at each iteration is
  determined from the metric field gradient taken at
  the midway point, i.e.
  $\nabla \Pi_{\sim} \left(I\circ\phi_1(0.5),J\circ\phi_2(0.5)\right)$.
  The full
  forward and inverse transforms are found through
  composition, i.e. $\phi=\phi_1 \circ \phi_2^{-1}$ and
  $\phi^{-1}=\phi_2 \circ \phi_1^{-1}$.
  }
  \label{fig:syn}
\end{figure}

@@

Although presenting a rigorous framework for image registration solutions
with desirable properties, the complexity of these diffeomorphic methodologies
requires substantial computational resources.  For typical 3-D neuroimaging
applications, the corresponding solutions require numerical integration over
and storage of 4-D velocity fields at each iteration which is limiting for
many common computational platforms.

Therefore, in addition to the full-scale
symmetric SyN offering described in \cite{avants2008}, the authors therein
provided a ``greedy'' alternative which has demonstrated superior performance
in various applications \citep{avants2011,klein2009,murphy2011}%
\footnote{
SyN was also the standard registration for the MICCAI 2013
Workshop on Segmentation:  Algorithms Theory, and Applications (SATA)
and corresponding challenge (https://masi.vuse.vanderbilt.edu/workshop2013).
}
while simultaneously being capable of running with limited computational
resources.  This is due to the restriction of the discrete, time-parameterized
velocity field samples to their respective
endpoints, i.e. the $v_i(\mathbf{x},t)$ are sampled at $t \in \{0,0.5,1\}$
implying simultaneous storage of only four transform vector fields
$\phi_1(\mathbf{x})$,
$\phi_1^{-1}(\mathbf{x})$, $\phi_2(\mathbf{x})$, and $\phi_2^{-1}(\mathbf{x})$
(cf Figure \ref{fig:syn}).  Since the greedy SyN framework is the focus of the evaluation,
the algorithmic steps are briefly sketched
in Algorithm \ref{alg:syn}.

@@

Although several velocity field regularization
operators have been proposed, many algorithmic instantiations
default to Gaussian smoothing due to its simplicity both in implementation
and complexity terms.  As mentioned earlier, 
a viable and practical alternative is the DMFFD 
approach based on B-splines for explicit regularization of vector fields.
As presented in \cite{tustison2009}, given the similarity metric
$\Pi_{\sim}$, the $d$-dimensional update field (i.e. preconditioned
gradient field), $\delta v_{i_1,\ldots,i_d}$, is calculated from
\begin{align}
\label{eq:dmffd}
  \delta v_{i_1,\ldots,i_d} = \frac{ \sum_{c=1}^{N_{\Omega}} \left( \frac{\partial \Pi_\sim}{\partial \mathbf{x}} \right)_c \prod_{j=1}^d B_{i_j}(x_j^c)  
  \cdot \frac{\prod_{j=1}^d B_{i_j}^2 (x_j^c)}
  {\sum_{k_1=1}^{r+1}\ldots\sum_{k_d=1}^{r+1} 
  \prod_{j=1}^d B_{k_j}^2 (x_j^c)} }
  {\sum_{c=1}^{N_{\Omega}} \prod_{j=1}^d B_{i_j}^2 (x_j^c)}
\end{align}
where the set of $B(\cdot)$ are the univariate B-spline basis functions for
separately modulating regularity in the solution for each parametric dimension,
$N_\Omega$ is the number of voxels in the reference image domain,
$r$ is the spline order in all dimensions,%
\footnote{
In terms of implementation, spline orders can be specified separately for each dimension but, for simplicity,
we only specify a single spline order.
}
and $\frac{\partial \Pi_\sim}{\partial \mathbf{x}}$ is the spatial 
similarity metric gradient at voxel $c$.

@@

antsApplyTransforms --dimensionality 3 \
                    --input ${moving_labels} \
                    --reference-image ${fixed} \
                    --output ${moving_warped_labels} \
                    --n NearestNeighbor \
                    --transform ${prefix}1Warp.nii.gz \
                    --transform ${prefix}0GenericAffine.mat \
                    --default-value 0
\end{lstlisting}
In this example, we first calculate an initial translation transform by aligning
the centers of (intensity) mass (although alignment based on other features is possible). 
The resulting transform is then used as input 
for determining an optimal rigid transform.  Serial propagation of the resulting
composite transform continues until all optimal transforms have been determined.
Optimization for each stage is determined by the specified general parameters 
including: smoothing and downsampling of fixed and moving images, convergence
criteria (including number of iterations per resolution level) and metric (or
metrics).  Note that any pair of images can be specified per metric per stage.

Although the
resulting transforms for each stage can be written to disk as output, the 
default output consists of a condensed set of transforms where compatible 
transforms have been composed to a single transform.  For example, in the
above command call, the initial translation, rigid, and affine transforms are
combined into a single generic affine transform file with the results of the
deformable transform consisting of discrete vector fields.  The output
transform files can then be applied using the {\tt antsApplyTransforms}
program which permits composition of any number of transform files with 
different interpolation schemes.   For both programs interpolation is never
performed more than once.

@@

\begin{figure}[htb]
  \centering
  \includegraphics[width=0.95\textwidth]{templates.jpg}
  \caption{Canonical views for each of the five cohort-specific templates generated using
  the ANTs tools as described
  in \cite{avants2010}.   The pseudo-geodesic transform between subjects is created from 
  the composition of transforms to/from the relevant template.
  }
  \label{fig:templates}
\end{figure}

@@

\begin{figure}[htb]
  \centering
  \includegraphics[width=0.75\textwidth]{pseudo_geodesic.jpg}
  \caption{Illustration of generating a pseudo-geodesic for any two subjects
  within the MGH10 cohort.  Once the transforms between the
  template and each subject are calculated, the mapping between any two
  subjects is found by composition of forward and inverse transforms.  For
  example, in the MGH data set, the pseudo-geodesic transform to
  map Subject g4 to Subject g7 is found by composing the forward
  transform from g4 to the template with the inverse transform from the
  template to g7 (green dashed lines).
  }
  \label{fig:pseudo_geodesic}
\end{figure}

Comparative evaluation of the two SyN registration approaches was performed 
within each cohort using a ``pseudo-geodesic'' approach.  Instead of
registering every subject to every other subject within a data set,
we generated the transforms from each subject to a cohort-specific
shape/intensity template.  These templates were built using the ANTs script {\tt antsMultivariateTemplateConstruction.sh}
which is a multivariate implementation of the work described in
\cite{avants2010}. Canonical views for each of the five templates
used for this study are given in Figure \ref{fig:templates}.  
Since calculation of the transform from each subject to the template also
includes generation of the corresponding inverse transform, the total
transformation from a given subject to any other is determined from the composition
of transforms mapping through the template.  An example illustration of the
geodesic approach is given in Figure \ref{fig:pseudo_geodesic}.

\begin{figure}[htb]
  \centering
  \includegraphics[width=0.95\textwidth]{labels.jpg}
  \caption{Axial views of sample labelings for a member of each data set.  
  The second row consists of the original labelings with the third row being
  refined versions of those labelings using the MALF algorithm \cite{wang2012}.
  These refinements provide more consistency between labelings and improved
  comparative assessments between algorithms.
  }
  \label{fig:labels}
\end{figure}

Additionally, we refined the labelings for each subject of each cohort
using the multi-atlas label fusion algorithm (MALF) developed by \cite{wang2012}
which is also distributed with ANTs.
For a given subject within a data set, every other subject was mapped to 
that subject using the pseudo-geodesic transform.  The set of transformed labelings 
were then used to determine a consensus
labeling for that subject.
This was to minimize the obvious {\it observer dimensionality artifacts} where manual
raters observe and label in a single dimension at a time.  This
is most easily seen in the axial or sagittal views of the different cohorts 
as labelings were done primarily in the coronal view (see Figure \ref{fig:labels}).  
We include both sets of results.  This provides two sets of labels per subject
for evaluation.

@@

\begin{figure}[htb]
  \centering
  \begin{tabular}{c}
  \includegraphics[width=0.95\textwidth]{evaluation.jpg}
  \end{tabular}
  \caption{Dice results for both algorithms for 
  each subject warped to every other subject using the pseudo-
  geodesic transform.  Each 
  row corresponds to one of the five data sets used for evaluation.
  For each data set we include a plotting of all individual label 
  Dice results by volume and a combined label box plotting.  The
  left and right halves show the respective results for 
  original and MALF-derived labelings.
  }
  \label{fig:dice}
\end{figure}

%\begin{figure}[htb]
%  \centering
%  \begin{tabular}{ccccc}
%  {} &
%  \multicolumn{2}{c}{\bf Original labeling} & 
%  \multicolumn{2}{c}{\bf MALF labeling} \\
%  \rotatebox{90}{\qquad\,\, CUMC12} &
%  \includegraphics[height=38mm]{versusCUMC.png} &
%  \includegraphics[height=38mm]{synComparisonCUMC.pdf} &
%  \includegraphics[height=38mm]{versusCUMCMALF.png} &
%  \includegraphics[height=38mm]{synComparisonCUMCMALF.pdf} \\
%  \rotatebox{90}{\qquad\quad IBSR18} &
%  \includegraphics[height=38mm]{versusIBSR.png} &
%  \includegraphics[height=38mm]{synComparisonIBSR.pdf} &
%  \includegraphics[height=38mm]{versusIBSRMALF.png} &
%  \includegraphics[height=38mm]{synComparisonIBSRMALF.pdf} \\
%  \rotatebox{90}{\qquad\quad LPBA40} &
%  \includegraphics[height=38mm]{versusLPBA.png} &
%  \includegraphics[height=38mm]{synComparisonLPBA.pdf} &
%  \includegraphics[height=38mm]{versusLPBAMALF.png} &
%  \includegraphics[height=38mm]{synComparisonLPBAMALF.pdf} \\
%  \rotatebox{90}{\qquad\quad MGH10} &
%  \includegraphics[height=38mm]{versusMGH.png} &
%  \includegraphics[height=38mm]{synComparisonMGH.pdf} &
%  \includegraphics[height=38mm]{versusMGHMALF.png} &
%  \includegraphics[height=38mm]{synComparisonMGHMALF.pdf} \\
%  \rotatebox{90}{\qquad\quad MAL35} &
%  \includegraphics[height=38mm]{versusMAL.png} &
%  \includegraphics[height=38mm]{synComparisonMAL.pdf} &
%  \includegraphics[height=38mm]{versusMALMALF.png} &
%  \includegraphics[height=38mm]{synComparisonMALMALF.pdf} \\
%  \end{tabular}
%  \caption{Dice results for both algorithms for 
%  each subject warped to every other subject using the pseudo-
%  geodesic transform.  Each 
%  row corresponds to one of the five data sets used for evaluation.
%  For each data set we include a plotting of all individual label 
%  Dice results by volume and a combined label box plotting.  The
%  left and right halves show the results for MALF-derived and 
%  original labelings, respectively.
%  }
%  \label{fig:dice}
%\end{figure}

>>>>

\selectlanguage{eng}
\begin{document}
\onecolumn
\firstpage{1}

\title[DMFFD Diffeomorphic Image Registration]{Explicit B-spline regularization in diffeomorphic image registration}
\author[\firstAuthorLast ]{\Authors}
\address{}
\correspondance{}
\editor{}
\topic{}

@@

\begin{abstract}
Diffeomorphic maps are central to image registration due largely to their
topological properties and success in providing biologically plausible
solutions to deformation and morphological
estimation problems. Popular variants of diffeomorphic algorithms include those characterized by
time-varying and constant velocity fields, and symmetrical considerations.
Prior information (i.e. regularization) is used to enforce transform plausibility taking the form of physics-based constraints or through some approximation thereof, e.g. Gaussian smoothing of the vector fields (a la Thirion's Demons \citep{thirion1998}).  In the context of the original Demons' framework, a free-form deformation methodological variant, the so-called {\it directly manipulated free-form deformation} \citep{tustison2009}, can be viewed as an alternative smoothing approach in which explicit regularization is achieved through 
fast B-spline appromixation.
This characterization can be used to provide alternative B-spline ``flavored'' diffeomorphic image registration solutions with several advantages which we describe in this work.  Implementation is open source and available through the Insight Toolkit and our Advanced Normalization Tools (ANTs) repository.  A thorough comparative evaluation with the well-known SyN algorithm \citep{avants2008}, implemented within the same framework, and its B-spline analog is performed using open data and open source evaluation tools.
\tiny
\keyFont{ \section{Keywords:}
open science, best practices, reproducibility, software }
\end{abstract}

@@

Establishment of anatomical and functional correspondence is a crucial
step towards gaining insight into biological processes.
Neuroscience research efforts, such as characterizing brain
morphology, require accurate and robust methods for producing such
mappings.  The extensive literature detailing methodology is evidence
of the rich history of algorithmic development which continues
contemporaneously.  We highlight several key historical contributions
which are particularly relevant to the work presented.

Free-form deformation (FFD) image registration, characterized by
regularization based on the B-spline basis functions, has several
advantages including algorithmic simplicity, good performance,
desirable properties, and several available implementations.  Current
research was preceded by related work for geometric modeling
\citep{sederberg1986} and originated with such important contributions
as \cite{szeliski1997,thevenaz1998,rueckert1999}.  Continued
development within this early spline-based paradigm produced
additional innovations such as integrated similarity metrics
\citep[e.g.][]{mattes2003}, additional transformation constraints
\citep[e.g.][]{rohlfing2003}, and notable open source implementations
\citep[e.g.][]{ibanez2005,klein2010,shackleford2010}.

@@

Continuing from the work of \cite{christensen1996} and subsequent
exploration into the mathematical formalisms of diffeomorphisms
\citep[e.g.][]{dupuis1998}, the well-known Large Deformation
Diffeomorphic Metric Mapping (LDDMM) algorithm was proposed in
\cite{beg2005}.  In contrast to the mapping produced by
\cite{christensen1996}, LDDMM yields the geodesic solution in the
space of diffeomorphisms between two images. Since its introduction,
LDDMM has inspired much innovation in the image registration
literature.  Applying the log-Euclidean framework of
\cite{arsigny2006}, DARTEL (Diffeomorphic Anatomic Registration using
Exponential Lie algebra) uses a constant velocity field
parameterization to provide a fast, diffeomorphic alternative
\citep{ashburner2007}. Additionally, symmetrical considerations in the
velocity field parameterization are discussed in \cite{avants2008} in
the context of a cross correlation similarity metric.  By explicitly
symmetrizing the LDDMM formulation, this Symmetric Normalization (SyN)
minimizes the bias of the resulting transformation when selecting the
``fixed'' and ``moving'' images.  A greedy version of this algorithm
has proven successful in neuroimaging \citep{klein2009} and pulmonary
\citep{murphy2011} applications as well as in multi-atlas label fusion \cite{wang2012}.

@@

Variants of three popular diffeomorphic algorithms and their DMFFD
analogs (LDDMM, DARTEL, and SyN) were implemented by the authors as
part of the recent refactoring of the open source Insight Toolkit
(ITK) although related work had been previously implemented within the
popular Advanced Normalization Tools (ANTs).%
\footnote{
http://stnava.github.io/ANTs
}
Given the popularity and excellent performance of the greedy variant
of SyN, our evaluation focus in this work is its B-spline
analog which we denote as ``B-spline SyN'' or ``DMFFD SyN.''
Evaluations of the respective algorithmic
instantiations are performed using the \verb#antsRegistration# program
found in the ANTs repository (also originally developed by the
authors).  This permits a direct algorithmic comparison as potential
sources for implementation bias have been reduced
\citep{tustison2013}.  Additionally, in the spirit of open science,
all text, figures, and scripts to reproduce the results contained in
this work are publicly available online.%
\footnote{
https://github.com/ntustison/BSplineMorphisms
}

@@

However,
as pointed out in \cite{avants2008}, implementations of this standard
LDDMM formulation are negatively affected by the lack of optimization
symmetry where arbitrary assignment of fixed and moving images could
lead to different solutions despite the fact that the theoretical
geodesic solution describes the same path forwards and backwards.
This observation led to the symmetric
formulation of Eq. (\ref{eq:lddmm}) found in \cite{avants2008}:
\begin{align}
  \inf_{\phi_1} \inf_{\phi_2} \left(
                     \int_0^{0.5} \left( \|v_1(t)\|_L^2 + \|v_2(t)\|_L^2 \right) dt +
                     \int_{\Omega} | I \circ \phi_1^{-1}(\mathbf{x},0.5)
                           - J \circ \phi_2^{-1}(\mathbf{x},0.5) |^2 d\Omega
              \right)
\end{align}
where
\begin{align}
  \frac{d \phi_i(\mathbf{x},t)}{dt} = v_i( \phi_i(\mathbf{x},t), t ),\,\, \phi_i(\mathbf{x},0) = \mathbf{Id}, \,\, i \in \{1,2\}
\end{align}
with extension to arbitrary similarity metric choice, i.e. the second
term is replaced with
\begin{align}
\int_{\Omega} \Pi_{\sim}
                          \left( I \circ \phi_1^{-1}(\mathbf{x},0.5),
                           J \circ \phi_2^{-1}(\mathbf{x},0.5) \right) d\Omega
\end{align}
with a popular choice for $\Pi_{\sim}$ being a local neighborhood cross
correlation \citep{avants2008,avants2011}.
Note that
$t$ is parameterized in opposite directions between $\phi_1$ and $\phi_2$.
A diagrammatic
illustration of the explicit symmetry associated with SyN is shown in
Figure \ref{fig:syn}.

\begin{figure}[htb]
  \centering
  \includegraphics[width=0.95\textwidth]{SyN.pdf}
  \caption{Illustration of the greedy SyN formulation.  Given images
  $I_A$ and $I_B$, the symmetric set-up requires finding the two
  transform pairs $\left(\phi_1,\phi_1^{-1}\right)$
  $\left(\phi_2,\phi_2^{-1}\right)$ which map to/from
  the respective images to the midway point. During
  optimization, the update field at each iteration is
  determined from the metric field gradient taken at
  the midway point, i.e.
  $\nabla \Pi_{\sim} \left(I\circ\phi_1(0.5),J\circ\phi_2(0.5)\right)$.
  The full
  forward and inverse transforms are found through
  composition, i.e. $\phi=\phi_1 \circ \phi_2^{-1}$ and
  $\phi^{-1}=\phi_2 \circ \phi_1^{-1}$.
  }
  \label{fig:syn}
\end{figure}

@@

Although presenting a rigorous framework for image registration solutions
with desirable properties, the complexity of these diffeomorphic methodologies
requires substantial computational resources.  For typical 3-D neuroimaging
applications, the corresponding solutions require numerical integration over
and storage of 4-D velocity fields at each iteration which is limiting for
many common, contemporary processing infrastructures.

Therefore, in addition to the full-scale
symmetric SyN offering described in \cite{avants2008}, the authors therein
provided a ``greedy'' alternative which has demonstrated superior performance
in various applications \citep{avants2011,klein2009,murphy2011}%
\footnote{
SyN was also the standard registration for the MICCAI 2013
Workshop on Segmentation:  Algorithms Theory, and Applications (SATA)
and corresponding challenge (https://masi.vuse.vanderbilt.edu/workshop2013).
}
while simultaneously being capable of running with limited computational
resources.  This is possible since the discrete time-parameterized
velocity field samples are restricted to their respective
endpoints, i.e. the $v_i(\mathbf{x},t)$ are sampled at $t \in \{0,0.5,1\}$
implying simultaneous storage of only four transform vector fields
$\phi_1(\mathbf{x})$,
$\phi_1^{-1}(\mathbf{x})$, $\phi_2(\mathbf{x})$, and $\phi_2^{-1}(\mathbf{x})$
(cf Figure \ref{fig:syn}).  Furthermore, the forward and inverse
mappings are guaranteed to be consistent within the discrete domain i.e. $\|
\phi_i^{-1}(\phi_i) - \mathbf{Id} \|^2 < \epsilon$.   
Since the greedy SyN framework is the focus of the evaluation,
the algorithmic steps are briefly sketched
in Algorithm \ref{alg:syn}.

@@

Although several velocity field regularization
operators have been proposed, many algorithmic instantiations
default to Gaussian smoothing due to its simplicity both in implementation
and complexity terms.  As mentioned earlier, 
a viable and practical alternative is the DMFFD 
approach based on B-splines for explicit regularization of vector fields.
As presented in \cite{tustison2009}, given the similarity metric
$\Pi_{\sim}$, the $d$-dimensional update field (i.e. preconditioned
gradient field), $\delta v_{i_1,\ldots,i_d}$, is calculated from
\begin{align}
\label{eq:dmffd}
  \delta v_{i_1,\ldots,i_d} = \frac{ \sum_{c=1}^{N_{\Omega}} \left( \frac{\partial \Pi_\sim}{\partial \mathbf{x}} \right)_c \prod_{j=1}^d B_{i_j}(x_j^c)  
  \cdot \frac{\prod_{j=1}^d B_{i_j}^2 (x_j^c)}
  {\sum_{k_1=1}^{r+1}\ldots\sum_{k_d=1}^{r+1} 
  \prod_{j=1}^d B_{k_j}^2 (x_j^c)} }
  {\sum_{c=1}^{N_{\Omega}} \prod_{j=1}^d B_{i_j}^2 (x_j^c)}
\end{align}
where the set of $B(\cdot)$ are the univariate B-spline basis functions for
separately modulating regularity in the solution for each parametric dimension,
$N_\Omega$ is the number of voxels in the reference image domain,
$r$ is the spline order in all dimensions,%
\footnote{
In terms of implementation spline orders can be specified separately for each dimension but, for simplicity,
we only specify a single spline order.
}
and $\frac{\partial \Pi_\sim}{\partial \mathbf{x}}$ is the spatial 
similarity metric gradient indexed by $c$.

@@

antsApplyTransforms --dimensionality 3 \
                    --input ${moving_labels} \
                    --reference-image ${fixed} \
                    --output ${moving_warped_labels} \
                    --n NearestNeighbor \
                    --transform ${prefix}1Warp.nii.gz \
                    --transform ${prefix}0GenericAffine.mat \
                    --default-value 0
\end{lstlisting}
In this example, we first calculate an initial translation transform by aligning
the centers of (intensity) mass. The resulting transform is then used as input 
for determining an optimal rigid transform.  Serial propagation of the resulting
composite transform continues until all optimal transforms have been determined.
Optimization for each stage is determined by the specified general parameters 
including: smoothing and downsampling of fixed and moving images, convergence
criteria (including number of iterations per resolution level) and metric (or
metrics).  Note that any pair of images can be specified per stage.

Although the
resulting transforms for each stage can be written to disk as output, the 
default output consists of a condensed set of transforms where compatible 
transforms have been composed to a single transform.  For example, in the
above command call, the initial translation, rigid, and affine transforms are
combined into a single generic affine transform file with the results of the
deformable transform consisting of discrete vector fields.  The output
transform files can then be applied using the {\tt antsApplyTransforms}
program which permits composition of any number of transform files with 
various interpolation schemes.   For both programs interpolation is never
performed more than once.

@@

\begin{figure}[htb]
  \centering
  \includegraphics[width=0.95\textwidth]{templates.pdf}
  \caption{Canonical views for each of the five cohort-specific templates generated using
  the ANTs tools as described
  in \cite{avants2010}.   The pseudo-geodesic transform between subjects is created from 
  the composition of transforms to/from the relevant template.
  }
  \label{fig:templates}
\end{figure}

@@

\begin{figure}[htb]
  \centering
  \includegraphics[width=0.75\textwidth]{pseudo_geodesic.pdf}
  \caption{Illustration of generating a pseudo-geodesic for any two subjects
  within the MGH10 cohort.  Once the transforms between the
  template and each subject are calculated, the mapping between any two
  subjects is found by composition of forward and inverse transforms.  For
  example, in the MGH data set, the pseudo-geodesic transform to
  map Subject g4 to Subject g7 is found by composing the forward
  transform from g4 to the template with the inverse transform from the
  template to g7 (green dashed lines).
  }
  \label{fig:pseudo_geodesic}
\end{figure}

Comparative evaluation of the two SyN registration approaches was performed 
within each cohort using a ``pseudo-geodesic'' approach.  Instead of
registering every subject to every other subject within a data set,
we generated the transforms from each subject to a cohort-specific
shape/intensity template.  This was accomplished using the ANTs script {\tt antsMultivariateTemplateConstruction.sh}
which is a multivariate implementation of the work described in
\cite{avants2010}. Canonical views for each of the five templates
used for this study are given in Figure \ref{fig:templates}.  
Since calculation of the transform from each subject to the template also
includes generation of the corresponding inverse transform, the total
transformation from a given subject to any other is determined from the composition
of transforms mapping through the template.  An example illustration of the
geodesic approach is given in Figure \ref{fig:pseudo_geodesic}.

\begin{figure}[htb]
  \centering
  \includegraphics[width=0.95\textwidth]{labels.pdf}
  \caption{Axial views of sample labelings for a member of each data set.  
  The second row consists of the original labelings with the third row being
  refined versions of those labelings using the MALF algorithm \cite{wang2012}.
  These refinements provide more consistency between labelings and improved
  comparative assessments between algorithms.
  }
  \label{fig:labels}
\end{figure}

Additionally, we refined the labelings for each subject of each cohort
using the multi-atlas label fusion algorithm (MALF) developed by \cite{wang2012}
which is also distributed with the ANTs repository.
For a given subject within a data set, every other subject was mapped to 
that subject using the pseudo-geodesic transform.  The set of transformed labelings 
were then used to determine a consensus
labeling for that subject.
This was to minimize the obvious {\it observer dimensionality artifacts} where manual
raters observe and label in a single dimension at a time.  This
is most easily seen in the axial or sagittal views of the different cohorts 
as labelings were done primarily in the coronal view (see Figure \ref{fig:labels}).  
We include both sets of results.  This provides two sets of labels per subject
for evaluation.

@@

\begin{figure}[htb]
  \centering
  \begin{tabular}{ccccc}
  {} &
  \multicolumn{2}{c}{\bf Original labeling} & 
  \multicolumn{2}{c}{\bf MALF labeling} \\
  \rotatebox{90}{\qquad\,\, CUMC12} &
  \includegraphics[height=38mm]{versusCUMC.png} &
  \includegraphics[height=38mm]{synComparisonCUMC.pdf} &
  \includegraphics[height=38mm]{versusCUMCMALF.png} &
  \includegraphics[height=38mm]{synComparisonCUMCMALF.pdf} \\
  \rotatebox{90}{\qquad\quad IBSR18} &
  \includegraphics[height=38mm]{versusIBSR.png} &
  \includegraphics[height=38mm]{synComparisonIBSR.pdf} &
  \includegraphics[height=38mm]{versusIBSRMALF.png} &
  \includegraphics[height=38mm]{synComparisonIBSRMALF.pdf} \\
  \rotatebox{90}{\qquad\quad LPBA40} &
  \includegraphics[height=38mm]{versusLPBA.png} &
  \includegraphics[height=38mm]{synComparisonLPBA.pdf} &
  \includegraphics[height=38mm]{versusLPBAMALF.png} &
  \includegraphics[height=38mm]{synComparisonLPBAMALF.pdf} \\
  \rotatebox{90}{\qquad\quad MGH10} &
  \includegraphics[height=38mm]{versusMGH.png} &
  \includegraphics[height=38mm]{synComparisonMGH.pdf} &
  \includegraphics[height=38mm]{versusMGHMALF.png} &
  \includegraphics[height=38mm]{synComparisonMGHMALF.pdf} \\
  \rotatebox{90}{\qquad\quad MAL35} &
  \includegraphics[height=38mm]{versusMAL.png} &
  \includegraphics[height=38mm]{synComparisonMAL.pdf} &
  \includegraphics[height=38mm]{versusMALMALF.png} &
  \includegraphics[height=38mm]{synComparisonMALMALF.pdf} \\
  \end{tabular}
  \caption{Dice results for both algorithms for 
  each subject warped to every other subject using the pseudo-
  geodesic transform.  Each 
  row corresponds to one of the five data sets used for evaluation.
  For each data set we include a plotting of all individual label 
  Dice results by volume and a combined label box plotting.  The
  left and right halves show the results for MALF-derived and 
  original labelings, respectively.
  }
  \label{fig:dice}
\end{figure}

As mentioned previously, a template was constructed for each data
set (cf Figure \ref{fig:templates}) from all cohort images.  Subsequently, 
each image was registered to its corresponding template using either 
SyN or B-spline SyN as described previously (prior linear registration 
stages were identical between the two algorithms).  As a brute-force parameter
exploration is not a part of this work, we rely on previously
reported research \citep{klein2009,avants2011} and our own experience as
authors/developers of the algorithm/software to select parameters which 
demonstrate robust performance across data sets.  For both algorithms,
the gradient step was 0.1 during each of the four multi-resolution levels
with shrink factors of $\{6,4,2,1\}$ with Gaussian smoothing for each
of those levels being $\mathcal{N}\left(0,\{9,4,1,0\}\right)$ in terms
of voxels.  The number of iterations per level were $\{100,100,70,20\}$
with a convergence threshold of $10^{-9}$ and window size of 15 iterations.

All processing was performed using the computational cluster at the 
University of Virginia.%
\footnote{
http://www.uvacse.virginia.edu
}
All perl scripts used to create the jobs for the cluster are
included in the github account associated with this evaluation.  
It should be noted that for the
data sets used in this study, times for B-spline SyN were approximately 
15-40\% greater than Gaussian-based SyN using single-threading and a dense
metric gradient sampling.
